---
title: "R Notebook"
output: html_notebook
---

Run Monocol 3 on 60 days midbrain organoids


```{r}

library(Seurat)
library(reticulate)
library(Matrix)
library(dplyr)
library(cowplot)
library(ggplot2)
library(dplyr)
library(DESeq2)
library(monocle3)
library(SeuratWrappers)




```


Read in the data

```{r}

outpath <- "/Users/rhalenathomas/Documents/Data/scRNAseq/AIWtrio60days/pseudotimeanalysis/"


seu.60 <- readRDS("/Users/rhalenathomas/Documents/Data/scRNAseq/AIWtrio60days/AWI002ParkinKOPinkKO60days_labels_14052022.rds")
DefaultAssay(seu.60) <- "RNA"


```


I'm going to subset out the control 60 days only

```{r}

Idents(seu.60) <- 'orig.ident'
unique(seu.60$orig.ident)

aiw <- subset(seu.60, idents = "Control")

DimPlot(aiw, group.by = 'cluster.ids')

# all the cell types are still there 



```


See which cluster were labelled as which cells in the 60days data

```{r}

#cell.key <- as.data.frame(table(seu.60$seurat_clusters,seu.60$cluster.ids)) 

t.lables <- as.data.frame(table(aiw$seurat_clusters,aiw$cluster.ids))
t.lables$Freq <- as.double(t.lables$Freq)

cell.key <-as.data.frame(t.lables  %>% group_by(Var1)  %>% top_n(1, Freq))
cell.key


```

Re-process the Seurat object

```{r}

aiw <- ScaleData(aiw,verbose = FALSE)
aiw <- FindVariableFeatures(aiw)
aiw<- RunPCA(aiw, npcs = 30, verbose = FALSE)
aiw <- RunUMAP(aiw, reduction = "pca", dims = 1:30)

DimPlot(aiw, reduction = "umap")
DimPlot(aiw, group.by = "cluster.ids")

```




Make a monocol3 object specifying each section


```{r}

# they added gene symbols for short names
# my short names are the same
# library("org.Hs.eg.db")
# gene_symbol <- as.list(org.Hs.egSYMBOL)

# ******* FIX GENE SHORT NAMES FOR LATER *********
# get the expression matrix from seurat object
expression_matrix <- GetAssayData(aiw, assay = "integrated", slot = "data")
raw_matrix <- GetAssayData(aiw, assay = "RNA", slot = "counts")

cell_info <- aiw@meta.data
gene_name <- raw_matrix@Dimnames[[1]]
gene_name <- sapply(gene_name, function(x) x[[1]][1])


# prepare for the monocle3 object
gene_name2 <- ifelse(is.na(gene_name), names(gene_name),gene_name)
gene_short_name <- gene_name2  # it is important to create this dataslot for later
gene_id <- rownames(raw_matrix)

genes_info <- cbind(gene_id, gene_short_name)
rownames(genes_info) <- rownames(raw_matrix)


cds <- new_cell_data_set(expression_data = raw_matrix, 
                         cell_metadata = cell_info, 
                         gene_metadata = genes_info)


cds 

# now the meta data is under colData 




```

Run the monocle UMAP 

```{r}

cds <- preprocess_cds(cds, method = "PCA", num_dim = 30, norm_method = "log")
plot_pc_variance_explained(cds)

# matching the conditions I used in seurat dim = 1:30, default knn = 30L
cds <- reduce_dimension(cds, reduction_method = "UMAP",umap.n_neighbors = 30L, preprocess_method = "PCA")
plot_cells(cds)


```

Look at some genes

```{r}

plot_cells(cds, genes = c("TH","S100B","MAP2","CALB1"))



```





Now we can plot coloured by a metadata

```{r}

plot_cells(cds, color_cells_by = "cluster.ids")
plot_cells(cds, color_cells_by = "cluster.ids",
           label_cell_groups = TRUE,label_principal_points = FALSE,
           label_branch_points=FALSE,
            show_trajectory_graph=FALSE,
           label_leaves=FALSE,
           group_label_size = 5)
plot_cells(cds, color_cells_by = "seurat_clusters")
plot_cells(cds, color_cells_by = "orig.ident")

colnames(cds@colData)

pdf(paste(outpath,"UMAPwithclusterids.pdf"), width = 10, height = 6)
plot_cells(cds, color_cells_by = "cluster.ids", 
           label_cell_groups = FALSE, label_branch_points = FALSE,
           label_principal_points = FALSE,
            show_trajectory_graph=FALSE,
           label_leaves=FALSE)
dev.off()


```


```{r}
# seurat UMAP

DimPlot(seu.60, group.by = 'cluster.ids')
DimPlot(seu.60, group.by = 'cluster.ids',split.by = 'orig.ident')




```


I will cluster with the monocol method and see if I'm happier with these cluster

```{r}
cds <- cluster_cells(cds, k = 75, resolution = 0.06)


plot_cells(cds)
plot_cells(cds, color_cells_by = "partition", group_cells_by = "partition") # partition is larger groups of the smaller clusters 
plot_cells(cds, color_cells_by = "cluster", group_cells_by = "cluster")

```

Can I get a table with my annotations by cluster id

```{r}

#table(cds@clusters$UMAP$clusters, cds@colData$cluster.ids)

t.lables <- as.data.frame(table(cds@clusters$UMAP$clusters, cds@colData$cluster.ids))
t.lables$Freq <- as.double(t.lables$Freq)

cell.key.2 <-as.data.frame(t.lables  %>% group_by(Var1)  %>% top_n(1, Freq))
cell.key.2

# much more similar to the seurat clusters


```

I will annotate the cell types from the previous annotation

```{r}
colData(cds)$assigned_cell_type <- as.character(clusters(cds))

```


```{r}

colData(cds)$assigned_cell_type <- dplyr::recode(colData(cds)$assigned_cell_type,
                                                 "1" = "Dopaminergic Neurons 1 and early DA",
                                                 "2" = "DA and other Neurons",
                                                 "3" = "RG 1a",
                                                 "4" = "RG 1b",
                                                 "5" = "RG 1c",
                                                 "6" = "RG 1d",
                                                 "7" = "RG 1e",
                                                 "8" = "Dopaminergic Neurons early TH",
                                                 "9" = "RG 1f",
                                                 "10" = "Dopaminergic Neurons 1 TH",
                                                 "11" = "RG 1g",
                                                 "12" = "NPC and RG",
                                                 "13" = "RG 1h",
                                                 "14" = "Astrocytes",
                                                 "15" = "RG 1i",
                                                 "16" = "RG and NPC",
                                                 "17" = "NPC-RG-earlyDA",
                                                 "18" = "Dopaminergic Neurons 2 and early DA",
                                                 "19" = "Neural Epithelial",
                                                 "20" = "NPCs",
                                                 "21" = "RG 2 astro",
                                                 "22" = "Dopaminergic Neurons early",
                                                 )
                                                 
plot_cells(cds, group_cells_by = "cluster", color_cells_by = "assigned_cell_type", label_cell_groups = TRUE)


```


Plot the new cluster labels 



```{r}
pdf(paste(outpath,"UMAPwithcellassignments.pdf"), width = 11, height = 6)
plot_cells(cds, color_cells_by = "assigned_cell_type", 
           label_cell_groups = FALSE, label_branch_points = FALSE,
           label_principal_points = FALSE,
            show_trajectory_graph=FALSE,
           label_leaves=FALSE)
dev.off()


plot_cells(cds, color_cells_by = "assigned_cell_type", 
           label_cell_groups = FALSE, label_branch_points = FALSE,
           label_principal_points = FALSE,
            show_trajectory_graph=FALSE,
           label_leaves=FALSE)

plot_cells(cds, color_cells_by = "assigned_cell_type", 
           label_cell_groups = TRUE, label_branch_points = FALSE,
           label_principal_points = FALSE,
            show_trajectory_graph=FALSE,
           label_leaves=FALSE, group_label_size = 5)


```






Now learn the trajectory graph

```{r}

cds <- learn_graph(cds)
plot_cells(cds,
           color_cells_by = "assigned_cell_type",
           label_groups_by_cluster=FALSE,
           label_leaves=FALSE,
           label_branch_points=FALSE)



plot_cells(cds,
           color_cells_by = "assigned_cell_type",
           label_leaves=FALSE,
           label_branch_points=TRUE,
           label_cell_groups = FALSE)



plot_cells(cds,
           color_cells_by = "assigned_cell_type",
           label_leaves=TRUE,
           label_branch_points=TRUE,
           label_cell_groups = FALSE,
           graph_label_size = 3)


#pdf(paste(outpath,"UMAPtrajectoryBranchLeaf.pdf"), width = 9, height = 3)
#plot_cells(cds,
         #  color_cells_by = "assigned_cell_type",
         #  label_leaves=TRUE,
         #  label_branch_points=TRUE,
         #  label_cell_groups = FALSE,
         #  graph_label_size = 3)
#dev.off()


```

This graph above will be used in downstream steps


It also seems the trajectory is set on top of the UMAP and thus the settings of the UMAP will change the output.
Lines are not connected across partions - I need to play with the clustering to change them



Now I will be choosing root nodes

I need to see the cell types again

```{r}

plot_cells(cds,
           color_cells_by = "assigned_cell_type",
           label_cell_groups=FALSE,
           label_leaves=TRUE,
           label_branch_points=TRUE,
           graph_label_size=3)
```



Save the monocole object 

```{r}
saveRDS(cds, paste(outpath,"CDSmonocle3_AIWcontrol.RDS",sep = ""))

```



```{r}
cds <- order_cells(cds)

```



Now I plot the cells coloured by pseudotime

```{r}
plot_cells(cds,
           color_cells_by = "pseudotime",
           label_cell_groups=FALSE,
           label_leaves=FALSE,
           label_branch_points=FALSE,
           graph_label_size=3)


pdf(paste(outpath,"PseudotimeUMAP.pdf", sep=""), width = 9, height = 5)
plot_cells(cds,
           color_cells_by = "pseudotime",
           label_cell_groups=FALSE,
           label_leaves=FALSE,
           label_branch_points=FALSE,
           graph_label_size=3)
dev.off()



```


Subset cells based on their trajectory

```{r}
cds_sub_TH <- choose_graph_segments(cds)

```


Working with 3D trajectories


```{r}
cds_3d <- reduce_dimension(cds, max_components = 3)
cds_3d <- cluster_cells(cds_3d)
cds_3d <- learn_graph(cds_3d)


cds_3d <- order_cells(cds_3d)


```


```{r}

cds_3d_plot_obj <- plot_cells_3d(cds_3d, color_cells_by="partition")
cds_3d_plot_obj

plot_cells_3d(cds_3d, color_cells_by="cluster.ids")

install.packages("htmlwidgets")
library("htmlwidgets")

saveWidget(widget = cds_3d_plot_obj,
        file = "/Users/rhalenathomas/Documents/Data/scRNAseq/AIWtrio60days/pseudotimeanalysis/pseudotime3DUMAP.html", 
                selfcontained = TRUE #creates a single html file
                )






```

Finding genes that change as a function of pseudotime

Once again, we turn to graph_test(), this time passing it neighbor_graph="principal_graph", which tells it to test whether cells at similar positions on the trajectory have correlated expression:

```{r}

cds_pr_test_res <- graph_test(cds, neighbor_graph="principal_graph", cores=4)
pr_deg_ids <- row.names(subset(cds_pr_test_res, q_value < 0.05))



```


Have a look at the experssion of some of the gene changing in pseudotime


```{r}

p1 <- plot_cells(cds, genes=c("STMN2", "DCX", "MKI67", "TPBG"),
           show_trajectory_graph=FALSE,
           label_cell_groups=FALSE,
           label_leaves=FALSE)

p2 <- plot_cells(cds, genes=c("INA", "NR2F1", "PBK", "NSG2"),
           show_trajectory_graph=FALSE,
           label_cell_groups=FALSE,
           label_leaves=FALSE)


p1
p2




plot_cells(cds, genes=c("EEF1A1","RPL41","RPS12","NRXN1"),
           show_trajectory_graph=FALSE,
           label_cell_groups=FALSE,
           label_leaves=FALSE)

plot_cells(cds, genes=c("PBK","VIM","UBE2C","IGFBP7"),
           show_trajectory_graph=FALSE,
           label_cell_groups=FALSE,
           label_leaves=FALSE)




```

Collect the genes into modules

```{r}

gene_module_df <- find_gene_modules(cds[pr_deg_ids,], resolution=c(10^seq(-6,-1)))



```



```{r}
cell_group_df <- tibble::tibble(cell=row.names(colData(cds)), 
                                cell_group=colData(cds)$assigned_cell_type)
agg_mat <- aggregate_gene_expression(cds, gene_module_df, cell_group_df)
row.names(agg_mat) <- stringr::str_c("Module ", row.names(agg_mat))

pheatmap::pheatmap(agg_mat,
                   scale="column", clustering_method="ward.D2")



pdf(paste(outpath,"celltypeGeneMod.pdf", sep = ""), width =6, height = 8)
pheatmap::pheatmap(agg_mat,
                   scale="column", clustering_method="ward.D2") + theme(axis.text.x = element_text(size = 50))

dev.off()



```


We can look at the modules and see which ones are associated with a specific cell type.

```{r}

# m 69, 72,64, 8, 30 asc with RG 1f also with oseom other RG /NPC groups
# mode 23 strong RG1 g and e

# mode 34 DA and other neurons
# RG 1h and b mod 15

# mode 12 DA early  and also in the TH neuron
# mode  78  DA early and RG 1i
# mode  51      RG1i

# mode 80 NPC also linked epi

# mod 38 astrocytes

# DA TH groups  mod 11, 48, 39 
# m 63 up in later TH neurons


# comparing different DA neurons

plot_cells(cds,
           genes=gene_module_df %>% filter(module %in% c(11, 48, 39, 63)),
           label_cell_groups=FALSE,
           show_trajectory_graph=FALSE)



plot_cells(cds,
           genes=gene_module_df %>% filter(module %in% c(8,23,51,38)),
           label_cell_groups=FALSE,
           show_trajectory_graph=FALSE)



```

Another way to see the genes 



Monocle offers another plotting function that can sometimes give a clearer view of a gene's dynamics along a single path. You can select a path with choose_cells() or by subsetting the cell data set by cluster, cell type, or other annotation that's restricted to the path. Let's pick one such path, the AFD cells:

```{r}


module23 <- gene_module_df %>% filter(module %in% 23)
# SLC38A11 ELOVL7  PCYT1B  KCNQ1

# other modules different in RG
mod15 <- gene_module_df %>% filter(module %in% 15)  # HES4 WNT4
mod55 <- gene_module_df %>% filter(module %in% 55)   # NEXN  LMX1A


cds_sub <- cds[rowData(cds)$gene_id %in% genesoi,
                       colData(cds)$assigned_cell_type %in% c("Dopaminergic Neurons early" )]

unique(cds@colData$assigned_cell_type)
#unique(cds@colData$cluster.ids)


# define subgroups 

da.sub <- c("Dopaminergic Neurons 2 and early DA", "Dopaminergic Neurons 1 TH", "Dopaminergic Neurons early TH",
            "DA and other Neurons","Dopaminergic Neurons 1 and early DA")
rg.sub <- c("RG 1a","RG 1b","RG 1c", "RG 1d","RG 1e","RG 1f", "RG 1g", "RG 1h", "RG 1i", "RG 1j",
            "NPC and RG","RG and NPC")

genesoi <- c("SLC38A11", "ELOVL7","PCYT1B", "KCNQ1")
genesoi <- c("TPBG")
genesoi <- c("WNT4")
genesoi <- c("VIM")

genesoi <- c("RTN1") # flat line

genesoi <- c("CD9")
cds_sub.cd9 <- cds[rowData(cds)$gene_id %in% genesoi,
                       colData(cds)$assigned_cell_type %in% rg.sub]
genesoi <- c("VIM")
cds_sub.vim <- cds[rowData(cds)$gene_id %in% genesoi,
                       colData(cds)$assigned_cell_type %in% rg.sub]

genesoi <- c("TOP2A")
genesoi <- c("SPARC")
cds_sub <- cds[rowData(cds)$gene_id %in% genesoi,
                       colData(cds)$assigned_cell_type %in% rg.sub]


genesoi <- c("DCX")
cds_sub.dcx <- cds[rowData(cds)$gene_id %in% genesoi,
                       colData(cds)$assigned_cell_type %in% da.sub]
genesoi <- c("STMN2")
cds_sub.stmn2 <- cds[rowData(cds)$gene_id %in% genesoi,
                       colData(cds)$assigned_cell_type %in% da.sub]
genesoi <- c("NR2F1")
cds_sub.nr2f1 <- cds[rowData(cds)$gene_id %in% genesoi,
                       colData(cds)$assigned_cell_type %in% da.sub]



```

The function plot_genes_in_pseudotime() takes a small set of genes and shows you their dynamics as a function of pseudotime:


```{r}

plot_genes_in_pseudotime(cds_sub.vim, color_cells_by = "assigned_cell_type", min_expr = 0.5) # vim
plot_genes_in_pseudotime(cds_sub.cd9, color_cells_by = "assigned_cell_type", min_expr = 0.5) # cd9
plot_genes_in_pseudotime(cds_sub, color_cells_by = "assigned_cell_type", min_expr = 0.5)

plot_genes_in_pseudotime(cds_sub.nr2f1, color_cells_by = "assigned_cell_type", min_expr = 0.5)
plot_genes_in_pseudotime(cds_sub.dcx, color_cells_by = "assigned_cell_type", min_expr = 0.5)
plot_genes_in_pseudotime(cds_sub.stmn2, color_cells_by = "assigned_cell_type", min_expr = 0.5)

# I believe the plots don't split by gene becasue the gene_short_name isn't added proporly 


```


Analyze subbranch trajectories

```{r}


cds_subset.rg <- choose_cells(cds)


```


And then calling graph_test() on the subset. This will identify genes with interesting patterns of expression that fall only within the region of the trajectory you selected, giving you a more refined and relevant set of genes.

```{r}

subset_pr_test_res.rg <- graph_test(cds_subset.rg, neighbor_graph="principal_graph", cores=4)
pr_deg_ids.rg <- row.names(subset(subset_pr_test_res.rg, q_value < 0.05))


```


Grouping these genes into modules can reveal fate specific genes or those that are activate immediate prior to or following the branch point:

```{r}
gene_module_df.rg <- find_gene_modules(cds_subset.rg[pr_deg_ids.rg,], resolution=0.001)


```


We will organize the modules by their similarity (using hclust) over the trajectory so it's a little easier to see which ones come on before others:

```{r}
agg_mat <- aggregate_gene_expression(cds_subset.rg, gene_module_df.rg)
module_dendro <- hclust(dist(agg_mat))
gene_module_df.rg$module <- factor(gene_module_df.rg$module, 
                                levels = row.names(agg_mat)[module_dendro$order])

plot_cells(cds_subset.rg,
           genes=gene_module_df.rg,
           label_cell_groups=FALSE,
           show_trajectory_graph=FALSE)


# check out some genes





```


Check for genes within modules

```{r}

mod1 <-gene_module_df.rg %>% filter(module %in% 1)
mod1.genes <- mod1[1:20,1]

mod1.gene.v <- as.vector(mod1.genes)



```

Look at these genes in pseudotime



```{r}


#cds_sub <- cds_subset.rg[rowData(cds_subset.rg)$gene_id %in% mod1.genes$id] # top 20 genes

#genesoi <- c("CCNL2")

g.list <- c("RERE","SPEN","PUM1")

for (g in g.list){
  cds_sub <- cds_subset.rg[rowData(cds_subset.rg)$gene_id %in% g]
  print(plot_genes_in_pseudotime(cds_sub, color_cells_by = "assigned_cell_type", min_expr = 0.5))
  print(g)
}


g = "RERE" # curve
g = "SPEN"
g = "PUM1"
cds_sub <- cds_subset.rg[rowData(cds_subset.rg)$gene_id %in% g]
  print(plot_genes_in_pseudotime(cds_sub, color_cells_by = "assigned_cell_type", min_expr = 0.5))


```


See these genes one the UMAP



```{r}
plot_cells(cds_subset.rg,
           genes=g.list,
           label_cell_groups=FALSE,
           show_trajectory_graph=FALSE)






```



What is we used a different root area?  
Choosing the dividing cells?

```{r}

plot_cells(cds, genes = c("MKI67","TOP2A"), label_cell_groups=FALSE,
           show_trajectory_graph=FALSE)

plot_cells(cds, genes = c("MKI67","TOP2A"),
           show_trajectory_graph=FALSE, color_cells_by = "assigned_cell_type", label_cell_groups = TRUE)





```

Have a look in the RG1 subset for ipsc or stem cell markers


```{r}

plot_cells(cds_subset.rg,
           genes= c("MKI67","TOP2A"),
           label_cell_groups=FALSE,
           show_trajectory_graph=FALSE)

plot_cells(cds_subset.rg,
           genes= c("TOP2A"),
           label_cell_groups=FALSE,
           show_trajectory_graph=TRUE)

plot_cells(cds_subset.rg,
           genes= c("TOP2A"),
           show_trajectory_graph=FALSE,color_cells_by = "assigned_cell_type", label_cell_groups = TRUE)


# check some other iPSC markers
# FUT4 is SSEA1

stem <- c("AP","CD9","FUT4","FUT2")
pluri <- c("POU5F1","SOX2","NANOG","KLF4")
# OCT4 = POU5F1
stem.p <- c("TRA1","SOX9","SOX2")
div <- c("MKI67","PCNA","TOP2A", "MYC")

plot_cells(cds_subset.rg,
           genes= stem,
           show_trajectory_graph=FALSE,color_cells_by = "assigned_cell_type", label_cell_groups = TRUE)


plot_cells(cds,
           genes= stem,
           show_trajectory_graph=FALSE,color_cells_by = "assigned_cell_type", label_cell_groups = TRUE)

plot_cells(cds,
           genes= stem,
           show_trajectory_graph=FALSE,color_cells_by = "assigned_cell_type", label_cell_groups = FALSE)

plot_cells(cds,
           genes= stem.p,
           show_trajectory_graph=FALSE,color_cells_by = "assigned_cell_type", label_cell_groups = FALSE)

plot_cells(cds,
           genes= pluri,
           show_trajectory_graph=FALSE,color_cells_by = "assigned_cell_type", label_cell_groups = FALSE)

plot_cells(cds,
           genes= div,
           show_trajectory_graph=FALSE,color_cells_by = "assigned_cell_type", label_cell_groups = FALSE)


```


Try pseudotime selecting different start points

```{r}

cds2 <- cds

cds2 <- order_cells(cds2)

# selecting RG1 f tip 


```



```{r}
plot_cells(cds2,
           color_cells_by = "pseudotime",
           label_cell_groups=FALSE,
           label_leaves=FALSE,
           label_branch_points=FALSE,
           graph_label_size=3)
```

```{r}

cds2 <- order_cells(cds2)
# choose RG1 f tip and RG2 as two start points and NPC and RG as another point


```

```{r}
plot_cells(cds2,
           color_cells_by = "pseudotime",
           label_cell_groups=FALSE,
           label_leaves=FALSE,
           label_branch_points=FALSE,
           graph_label_size=3)
```

This is more sensible in trying to align pseudo time with developmental time
Check some of the other factors.

```{r}
cds_pr_test_res2 <- graph_test(cds2, neighbor_graph="principal_graph", cores=4)
pr_deg_ids2 <- row.names(subset(cds_pr_test_res2, q_value < 0.05))

```


See if the same genes are identified

```{r}

plot_cells(cds, genes=c("EEF1A1","RPL41","RPS12","NRXN1"),
           show_trajectory_graph=FALSE,
           label_cell_groups=FALSE,
           label_leaves=FALSE)

```








Add the monocle cell ids into the seurat object


```{r}


aiw <- AddMetaData(aiw, cds@colData$assigned_cell_type, col.name = "mon.cell.ids")

saveRDS(aiw, "/Users/rhalenathomas/Documents/Data/scRNAseq/AIWtrio60days/AIWcontrolNov28.RDS")
 

```









