---
title: "R Notebook"
output: html_notebook
---

Run Monocol 3 on 60 days midbrain organoids


```{r}

rm(list = ls(all.names = TRUE))
library(Seurat)
library(reticulate)
library(Matrix)
library(dplyr)
library(cowplot)
library(ggplot2)
library(dplyr)
library(DESeq2)
library(monocle3)
library(SeuratWrappers)
library(SeuratDisk)
library(rhdf5)



```


Read in the data

Here we have a python object from scanpy the anndata objects h5ad and need to read it into R.

We use the library SeuratDisk and a two step process provided by Seurat.

```{r}

# set the directory to the input
inpath <- "/Users/rhalenathomas/Documents/Data/scRNAseq/PublicData/"
Convert("/Users/rhalenathomas/Documents/Data/scRNAseq/PublicData/DG_anndata.h5ad",dest = "h5seurat", overwrite = TRUE, assay = "RNA")

seu <- LoadH5Seurat("/Users/rhalenathomas/Documents/Data/scRNAseq/PublicData/DG_anndata.h5seurat",meta.data = FALSE, misc = FALSE)
# works but no cluster annotation

# need to fetch the metadata

# see what is in the h5
h5ls("/Users/rhalenathomas/Documents/Data/scRNAseq/PublicData/DG_anndata.h5ad")

# a list of 2 
temp <- h5read("/Users/rhalenathomas/Documents/Data/scRNAseq/PublicData/DG_anndata.h5ad", "/obs/clusters")

# the format is the level and then a code for each cell.  

```

```{r}

# how to make my meta data list into one vector
# make a df of the codes 
decode <- temp[[1]]
decode 
df.code <- as.data.frame(temp[[2]])
colnames(df.code) <- "code"
# decoding funciton

unique(temp[[2]])

df.code <- df.code %>% mutate(Cluster = case_when(code == 0 ~ "Astrocytes",
                                                  code == 1 ~ "Cajal Retzius",
                                                  code == 2 ~ "Cck-Tox",
                                                  code == 3 ~ "Endothelial",
                                                  code == 4 ~ "GABA",
                                                  code == 5 ~ "Granule immature",
                                                  code == 6 ~ "Granule mature",
                                                  code == 7 ~ "Microglia" ,
                                                  code == 8 ~ "Mossy",
                                                  code == 9 ~ "Neuroblast",
                                                  code == 10 ~ "OL",
                                                  code == 11 ~ "OPC",
                                                  code == 12 ~ "Radial Glia-like",
                                                  code == 13 ~ "nIPC"))


# now I can add this into the seurat object

seu <- AddMetaData(seu,metadata = df.code$Cluster, col.name = "Cluster")


```




Lets have a look at the DG data

```{r}
colnames(seu@meta.data)
# no meta data for cell types
DimPlot(seu)
Idents(seu) <- 'Cluster'
DimPlot(seu, label = TRUE)

```

Check some expression

```{r}
VlnPlot(seu, group.by = 'Cluster', features = c("Grin1","Gria2"))
# mouse genes

```


Now we need to make the seurat object in to a monocle 3 object


```{r}

# get the expression matrix from seurat object
expression_matrix <- GetAssayData(seu, slot = "data")
raw_matrix <- GetAssayData(seu, assay = "RNA", slot = "counts")

#expression_matrix <- seu@assays$RNA@scale.data
#raw_matrix <- seu@assays$RNA@counts

cell_info <- seu@meta.data
gene_name <- raw_matrix@Dimnames[[1]]

gene_name <- sapply(gene_name, function(x) x[[1]][1])


# prepare for the monocle3 object
gene_name <- ifelse(is.na(gene_name), names(gene_name),gene_name)
gene_short_name <- gene_name  # it is important to create this dataslot for later
gene_id <- rownames(raw_matrix)

genes_info <- cbind(gene_id, gene_short_name)
rownames(genes_info) <- rownames(raw_matrix)

cds <- new_cell_data_set(expression_data = raw_matrix, 
                         cell_metadata = cell_info, 
                         gene_metadata = genes_info)

cds 

# now the meta data is under colData 



```

Check out the cds object

```{r}

unique(cds@colData$Cluster)

```
Run the steps to make the UMAP

```{r}

# default conditions

cds <- preprocess_cds(cds, method = "PCA", norm_method = "log")
plot_pc_variance_explained(cds)


cds <- reduce_dimension(cds, reduction_method = "UMAP", preprocess_method = "PCA")
plot_cells(cds)


```

See the UMAP and clusters with the monocle 3 functions


```{r}

plot_cells(cds, color_cells_by = "Cluster") # labels I added 
plot_cells(cds, color_cells_by = "Cluster",
           label_cell_groups = TRUE,label_principal_points = FALSE,
           label_branch_points=FALSE,
            show_trajectory_graph=FALSE,
           label_leaves=FALSE,
           group_label_size = 5)



pdf(paste(outpath,"UMAPwithclusterids.pdf"), width = 10, height = 6)
plot_cells(cds, color_cells_by = "cluster.ids", 
           label_cell_groups = FALSE, label_branch_points = FALSE,
           label_principal_points = FALSE,
            show_trajectory_graph=FALSE,
           label_leaves=FALSE)
dev.off()


```


Need to cluster

```{r}
cds <- cluster_cells(cds) # default settings


plot_cells(cds, color_cells_by = "partition", group_cells_by = "partition", group_label_size = 4) # partition is larger groups of the smaller clusters 
plot_cells(cds, color_cells_by = "cluster", group_cells_by = "cluster",
           group_label_size = 4)

```



Now learn the trajectory graph

```{r}

#cds <- learn_graph(cds)
plot_cells(cds,
           color_cells_by = "Cluster",
           label_groups_by_cluster=FALSE,
           label_leaves=FALSE,
           label_branch_points=FALSE,
           group_label_size = 4)


plot_cells(cds,
           color_cells_by = "Cluster",
           label_groups_by_cluster=FALSE,
           label_cell_groups = FALSE,
           label_leaves=FALSE,
           label_branch_points=TRUE,
           group_label_size = 8)

plot_cells(cds,
           color_cells_by = "Cluster",
           label_groups_by_cluster=FALSE,
           label_cell_groups = FALSE,
           label_leaves=TRUE,
           label_branch_points=FALSE,
           group_label_size = 8)


```

This graph above will be used in downstream steps


It also seems the trajectory is set on top of the UMAP and thus the settings of the UMAP will change the output.

```{r}

cds <- order_cells(cds)

```



Now I plot the cells coloured by pseudotime

```{r}
plot_cells(cds,
           color_cells_by = "pseudotime",
           label_cell_groups=FALSE,
           label_leaves=FALSE,
           label_branch_points=FALSE,
           graph_label_size=3)

```


Finding genes that change as a function of pseudotime

Once again, we turn to graph_test(), this time passing it neighbor_graph="principal_graph", which tells it to test whether cells at similar positions on the trajectory have correlated expression:

```{r}

cds_pr_test_res <- graph_test(cds, neighbor_graph="principal_graph", cores=4)
pr_deg_ids <- row.names(subset(cds_pr_test_res, q_value < 0.05))



```


Have a look at the experssion of some of the gene changing in pseudotime


```{r}

p1 <- plot_cells(cds, genes=c("Clqb", "Clqc", "Mog", "Clqa"),
           show_trajectory_graph=FALSE,
           label_cell_groups=FALSE,
           label_leaves=FALSE)

p2 <- plot_cells(cds, genes=c("Opalin", "Tyrobp", "Ctss", "Acta2"),
           show_trajectory_graph=FALSE,
           label_cell_groups=FALSE,
           label_leaves=FALSE)


p1
p2





```

Collect the genes into modules

```{r}

gene_module_df <- find_gene_modules(cds[pr_deg_ids,], resolution=c(10^seq(-6,-1)))



```



```{r}
cell_group_df <- tibble::tibble(cell=row.names(colData(cds)), 
                                cell_group=colData(cds)$Cluster)
agg_mat <- aggregate_gene_expression(cds, gene_module_df, cell_group_df)
row.names(agg_mat) <- stringr::str_c("Module ", row.names(agg_mat))

pheatmap::pheatmap(agg_mat,
                   scale="column", clustering_method="ward.D2")



pdf("/Users/rhalenathomas/Documents/Data/scRNAseq/PublicData/GeneModheatmap.pdf", width =6, height = 10)
pheatmap::pheatmap(agg_mat,
                  scale="column", clustering_method="ward.D2") + theme(axis.text.x = element_text(size = 50))
dev.off()

png("/Users/rhalenathomas/Documents/Data/scRNAseq/PublicData/GeneModheatmap.png", width =500, height = 600)
pheatmap::pheatmap(agg_mat,
                  scale="column", clustering_method="ward.D2") + theme(axis.text.x = element_text(size = 50))
dev.off()




```

We can look at the modules and see which ones are associated with a specific cell type.

```{r}

# m 6 mature granual cells
# m 5 neuroblasts
# m 11 and 22 iPC
# mod 3 and 53 astrocytes
# endothelial 54, 14, a lot
# microglia mod 50

plot_cells(cds,
           genes=gene_module_df %>% filter(module %in% c(6, 5, 3, 22)),
           label_cell_groups=FALSE,
           show_trajectory_graph=FALSE)



plot_cells(cds,
           genes=gene_module_df %>% filter(module %in% c(11,53,54,50)),
           label_cell_groups=FALSE,
           show_trajectory_graph=FALSE)



```

Another way to see the genes 



Monocle offers another plotting function that can sometimes give a clearer view of a gene's dynamics along a single path. You can select a path with choose_cells() or by subsetting the cell data set by cluster, cell type, or other annotation that's restricted to the path. Let's pick one such path, the AFD cells:

```{r}

# mature granule cells 
module6 <- gene_module_df %>% filter(module %in% 6)
# testing for genes with different slopes
# "Itm2c","Fam174a"

genesoi <- c("Itm2c","Fam174a","Scg2","Dock10")

cds_sub <- cds[rowData(cds)$gene_id %in% genesoi]

plot_genes_in_pseudotime(cds_sub, color_cells_by = "Cluster", min_expr = 0.5)

cds_sub <- cds[rowData(cds)$gene_id %in% genesoi, colData(cds)$Cluster %in% c("Granule mature","Granule immature","Neuroblast")]


plot_genes_in_pseudotime(cds_sub, color_cells_by = "Cluster", min_expr = 0.5)

```


Try another module

```{r}

# mature granule cells 
#module22 <- gene_module_df %>% filter(module %in% 22)
# testing for genes with different slopes
# "Sf3b1","Mgst3","Ndufs2"

genesoi <- c("Sf3b1","Mgst3","Ppa1","Ndufs2")

cds_sub <- cds[rowData(cds)$gene_id %in% genesoi]

plot_genes_in_pseudotime(cds_sub, color_cells_by = "Cluster", min_expr = 0.5)

cds_sub <- cds[rowData(cds)$gene_id %in% genesoi, colData(cds)$Cluster %in% c("Granule mature","Granule immature","Neuroblast","nIPC")]


plot_genes_in_pseudotime(cds_sub, color_cells_by = "Cluster", min_expr = 0.5)

cds_sub <- cds[rowData(cds)$gene_id %in% genesoi, colData(cds)$Cluster %in% c("Neuroblast")]


plot_genes_in_pseudotime(cds_sub, color_cells_by = "Cluster", min_expr = 0.5)



```






Analyze subbranch trajectories

```{r}


cds_subset <- choose_cells(cds)


```


And then calling graph_test() on the subset. This will identify genes with interesting patterns of expression that fall only within the region of the trajectory you selected, giving you a more refined and relevant set of genes.

```{r}

subset_pr_test_res.rg <- graph_test(cds_subset.rg, neighbor_graph="principal_graph", cores=4)
pr_deg_ids.rg <- row.names(subset(subset_pr_test_res.rg, q_value < 0.05))


```


Grouping these genes into modules can reveal fate specific genes or those that are activate immediate prior to or following the branch point:

```{r}
gene_module_df.rg <- find_gene_modules(cds_subset.rg[pr_deg_ids.rg,], resolution=0.001)


```


We will organize the modules by their similarity (using hclust) over the trajectory so it's a little easier to see which ones come on before others:

```{r}
agg_mat <- aggregate_gene_expression(cds_subset.rg, gene_module_df.rg)
module_dendro <- hclust(dist(agg_mat))
gene_module_df.rg$module <- factor(gene_module_df.rg$module, 
                                levels = row.names(agg_mat)[module_dendro$order])

plot_cells(cds_subset.rg,
           genes=gene_module_df.rg,
           label_cell_groups=FALSE,
           show_trajectory_graph=FALSE)


# check out some genes





```


Check for genes within modules

```{r}

mod1 <-gene_module_df.rg %>% filter(module %in% 1)
mod1.genes <- mod1[1:20,1]

mod1.gene.v <- as.vector(mod1.genes)



```

Look at these genes in pseudotime



```{r}


#cds_sub <- cds_subset.rg[rowData(cds_subset.rg)$gene_id %in% mod1.genes$id] # top 20 genes

#genesoi <- c("CCNL2")

g.list <- c("RERE","SPEN","PUM1")

for (g in g.list){
  cds_sub <- cds_subset.rg[rowData(cds_subset.rg)$gene_id %in% g]
  print(plot_genes_in_pseudotime(cds_sub, color_cells_by = "assigned_cell_type", min_expr = 0.5))
  print(g)
}


g = "RERE" # curve
g = "SPEN"
g = "PUM1"
cds_sub <- cds_subset.rg[rowData(cds_subset.rg)$gene_id %in% g]
  print(plot_genes_in_pseudotime(cds_sub, color_cells_by = "assigned_cell_type", min_expr = 0.5))


```


See these genes one the UMAP



```{r}
plot_cells(cds_subset.rg,
           genes=g.list,
           label_cell_groups=FALSE,
           show_trajectory_graph=FALSE)






```



What is we used a different root area?  
Choosing the dividing cells?

```{r}

plot_cells(cds, genes = c("MKI67","TOP2A"), label_cell_groups=FALSE,
           show_trajectory_graph=FALSE)

plot_cells(cds, genes = c("MKI67","TOP2A"),
           show_trajectory_graph=FALSE, color_cells_by = "assigned_cell_type", label_cell_groups = TRUE)





```

Have a look in the RG1 subset for ipsc or stem cell markers


```{r}

plot_cells(cds_subset.rg,
           genes= c("MKI67","TOP2A"),
           label_cell_groups=FALSE,
           show_trajectory_graph=FALSE)

plot_cells(cds_subset.rg,
           genes= c("TOP2A"),
           label_cell_groups=FALSE,
           show_trajectory_graph=TRUE)

plot_cells(cds_subset.rg,
           genes= c("TOP2A"),
           show_trajectory_graph=FALSE,color_cells_by = "assigned_cell_type", label_cell_groups = TRUE)


# check some other iPSC markers
# FUT4 is SSEA1

stem <- c("AP","CD9","FUT4","FUT2")
pluri <- c("POU5F1","SOX2","NANOG","KLF4")
# OCT4 = POU5F1
stem.p <- c("TRA1","SOX9","SOX2")
div <- c("MKI67","PCNA","TOP2A", "MYC")

plot_cells(cds_subset.rg,
           genes= stem,
           show_trajectory_graph=FALSE,color_cells_by = "assigned_cell_type", label_cell_groups = TRUE)


plot_cells(cds,
           genes= stem,
           show_trajectory_graph=FALSE,color_cells_by = "assigned_cell_type", label_cell_groups = TRUE)

plot_cells(cds,
           genes= stem,
           show_trajectory_graph=FALSE,color_cells_by = "assigned_cell_type", label_cell_groups = FALSE)

plot_cells(cds,
           genes= stem.p,
           show_trajectory_graph=FALSE,color_cells_by = "assigned_cell_type", label_cell_groups = FALSE)

plot_cells(cds,
           genes= pluri,
           show_trajectory_graph=FALSE,color_cells_by = "assigned_cell_type", label_cell_groups = FALSE)

plot_cells(cds,
           genes= div,
           show_trajectory_graph=FALSE,color_cells_by = "assigned_cell_type", label_cell_groups = FALSE)


```


Try pseudotime selecting different start points

```{r}

cds2 <- cds

cds2 <- order_cells(cds2)

# selecting RG1 f tip 


```



```{r}
plot_cells(cds2,
           color_cells_by = "pseudotime",
           label_cell_groups=FALSE,
           label_leaves=FALSE,
           label_branch_points=FALSE,
           graph_label_size=3)
```

```{r}

cds2 <- order_cells(cds2)
# choose RG1 f tip and RG2 as two start points and NPC and RG as another point


```

```{r}
plot_cells(cds2,
           color_cells_by = "pseudotime",
           label_cell_groups=FALSE,
           label_leaves=FALSE,
           label_branch_points=FALSE,
           graph_label_size=3)
```

This is more sensible in trying to align pseudo time with developmental time
Check some of the other factors.

```{r}
cds_pr_test_res2 <- graph_test(cds2, neighbor_graph="principal_graph", cores=4)
pr_deg_ids2 <- row.names(subset(cds_pr_test_res2, q_value < 0.05))

```


See if the same genes are identified

```{r}

plot_cells(cds, genes=c("EEF1A1","RPL41","RPS12","NRXN1"),
           show_trajectory_graph=FALSE,
           label_cell_groups=FALSE,
           label_leaves=FALSE)

```








Add the monocle cell ids into the seurat object


```{r}


aiw <- AddMetaData(aiw, cds@colData$assigned_cell_type, col.name = "mon.cell.ids")

saveRDS(aiw, "/Users/rhalenathomas/Documents/Data/scRNAseq/AIWtrio60days/AIWcontrolNov28.RDS")
 

```









